Topology and experimental distinguishability

### 2017-07-13 Articles/TopologyAndExperimentalDistinguishability/article.tex
Changed author ordering and corresponding author
Title, abstract, intro, and conclusion edits

\title{Topology and physical distinguishability}

%% Group authors per affiliation:
\author[]{Gabriele Carcassi\corref{cor1}}
\ead{carcassi@umich.edu}

\author[]{Christine A. Aidala}
\ead{caidala@umich.edu}

@@

\begin{abstract}
We formalize the relationship between topological spaces and the ability to distinguish objects experimentally, providing understanding and justification as to why topological spaces and continuous functions are pervasive tools in the physical sciences. The aim is to use these ideas as a stepping stone to give a more rigorous physical foundation to dynamical systems and, in particular, Hamiltonian dynamics.  We first define an experimental observation as a statement that can be verified using an experimental procedure and show that observations are closed under finite conjunction and countable disjunction. We then consider observations that identify elements in a set and show how they induce a Hausdorff and second-countable topology on that set, thus identifying an open set as one that can be associated with an experimental observation. We then show that only continuous functions preserve experimental distinguishability and that the collection of these functions can be given a Hausdorff and second-countable topology.

@@

\begin{keyword}
topological spaces \sep other keyword
\MSC[2010] 00-01\sep  99-00
\end{keyword}

@@

The successful use of mathematical ideas in experimental sciences is something long established and celebrated.\footnote{Unreasonable effectiveness and other} In particular topology is arguably the most widespread, as it provides the foundation for other mathematical tools such as differential geometry and ???. Why is it so successful? What physical property is captured by topological spaces that is so fundamental for scientific investigation?

We believe we have identified the answer: topologies capture the way objects are experimentally distinguishable. It is open sets that are associated with experimental outcomes, not the values themselves. For example, as real values are only measurable with finite precision, the standard topology on Euclidean spaces is the one that is physically meaningful. Along the same line, experimental relationships need to preserve what can be experimentally distinguished, and they are therefore represented as continuous functions.

@@

We believe this work not only provides interesting insights, but a firm justification of the use of topological ideas in the physical sciences.

>>>>

\title{Topology and experimental distinguishability}

%% Group authors per affiliation:
\author[]{Christine A. Aidala\corref{cor1}}
\ead{caidala@umich.edu}

\author[]{Gabriele Carcassi}
\ead{carcassi@umich.edu}

@@

\begin{abstract}
We formalize the relationship between topological spaces and the ability to distinguish objects experimentally, providing understanding and justification as to why topological spaces and continuous functions are pervasive tools in science. We first define an experimental observation as a statement that can be verified using an experimental procedure and show that observations are closed under finite conjunction and countable disjunction. We then consider observations that identify elements in a set and show how they induce a Hausdorff and second-countable topology on that set, thus identifying an open set as one that can be associated with an experimental observation. We then show that only continuous functions preserve experimental distinguishability and that the collection of these functions can be given a Hausdorff and second-countable topology.

@@

\begin{keyword}
topological spaces \sep TODO
\MSC[2010] TODO \sep TODO
\end{keyword}

@@

The successful use of mathematical ideas in experimental sciences is something long established and celebrated.\footnote{Unreasonable effectiveness and other} In particular topology is arguably the most widespread, as it provides the foundation for other mathematical tools such as differential geometry and ???. Why is it so successful? What property is captured by topological spaces that is so fundamental for scientific investigation?

We believe we have identified the answer: topologies capture the way objects are experimentally distinguishable. It is open sets that are associated with experimental outcomes, not the points themselves. For example, as real values are only measurable with finite precision, the standard topology on Euclidean spaces is the one that is physically meaningful. Along the same line, experimental relationships need to preserve what can be experimentally distinguished, and they are therefore represented as continuous functions.

@@

We believe this work not only provides interesting insights, but a firm justification of the use of topological ideas in the sciences.
